---
#############################################################
#                                                           #
# Click on "Run Document" in RStudio to run this worksheet. #
#                                                           #
#############################################################
title: "Aproximate Bayesian Computation"
author: "Miguel de Navascués"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library("phyclust", quietly=T)
source("readms.output.R", local =T)
Orange         <- "#E69F00"
Sky_Blue       <- "#56B4E9" 
Bluish_Green   <- "#009E73"
Yellow         <- "#F0E442"
Blue           <- "#0072B2"
Vermillion     <- "#D55E00"
Reddish_Purple <- "#CC79A7"
Black_transparency         <- rgb(  0,  0,  0,100,maxColorValue=255)
Orange_transparency        <- rgb(230,159,  0,100,maxColorValue=255)
SkyBlue_transparency       <- rgb( 86,180,233,100,maxColorValue=255)
BluishGreen_transparency   <- rgb(  0,158,115,100,maxColorValue=255)
Yellow_transparency        <- rgb(240,228, 66,100,maxColorValue=255)
Blue_transparency          <- rgb(  0,114,178,100,maxColorValue=255)
Vermillion_transparency    <- rgb(213, 94,  0,100,maxColorValue=255)
ReddishPurple_transparency <- rgb(204,121,167,100,maxColorValue=255)
```

## Setup

[Colorblind friendly palette](http://jfly.iam.u-tokyo.ac.jp/color/):

```{r color_setup, include=TRUE}
Orange         <- "#E69F00"
Sky_Blue       <- "#56B4E9" 
Bluish_Green   <- "#009E73"
Yellow         <- "#F0E442"
Blue           <- "#0072B2"
Vermillion     <- "#D55E00"
Reddish_Purple <- "#CC79A7"
```



## Approximation 1

Approximate Bayesian Computation is a likelihood-free model-based inference approach. The idea behind likelihood-free methods is that the analytical/numerical calculating of the likelihood is substituted by an estimate of the likelihood based on simulations.

### Exercise 1

For the coin toss experiment, estimate the likelihood of $p=0.5$ for $h=4$ and $t=6$ using simulations:

```{r simulation_coin_toss_rbinom, exercise=TRUE}
h = 4; t = 6; p = 0.5
n <- h+t
num_of_sim = _____
h_prime <- rbinom(n=num_of_sim,size=n,prob=p)
_____
```

```{r simulation_coin_toss_rbinom-hint}
h = 4; t = 6; p = 0.5
n <- h+t
num_of_sim = 100
h_prime <- rbinom(n=num_of_sim,size=n,prob=p)
_____(h_prime==h)/_____
```

```{r simulation_coin_toss_rbinom-solution}
h = 4; t = 6; p = 0.5
n <- h+t
num_of_sim = 100
h_prime <- rbinom(n=num_of_sim,size=n,prob=p)
sum(h_prime==h)/num_of_sim
```

The true value (for $p=0.5$, $h=4$ and $t=6$) is `r choose(10, 4) * 0.5^10`. Is the estimate reasonable? How can the estimate be improved?


### Exercise 2

Plot the likelihood profile from using simulation-based estimates of the likelihood. Compare the results with the analytical solution. Why should we bother about a (slower, less accurate) simulation-based solution? For this exercise you have available two functions: `flip_coin_likelihood()` and `flip_coin_likelihood_approx()` (see source code for details).

```{r likelihood_function-setup}
flip_coin_likelihood <- function(p,h,t){
  likelihood <- choose(h+t,h) * p^h * (1-p)^t
  return (list(p=p,l=likelihood))
}
flip_coin_likelihood_approx <- function(p,h,t,num_of_sim){
  likelihood <- array(dim=length(p))
  for (i in seq_along(p)){
    likelihood[i] <- sum(rbinom(n=num_of_sim,size=h+t,prob=p[i])==h)/num_of_sim  
  }
  return (list(p=p,l=likelihood))
}
```


```{r likelihood_function, exercise=TRUE}
h = 4; t = 6; num_of_sim = 1000
p = seq(0,1,0.01)
l_profile_true   <- flip_coin_likelihood(p,h,t)
l_profile_approx <- flip_coin_likelihood_approx(p,h,t,num_of_sim)
plot(x = l_profile_approx$p,
     y = l_profile_approx$l,
     xlab = "p", ylab = "Likelihood", type = "l", col = Bluish_Green, lwd=2)
lines(x = l_profile_true$p,
      y = l_profile_true$l,
      col = Bluish_Green)
```


Posterior probabilities can be also estimated through the use of simulations using the **rejection algorithm**. The following procedure produces a sample from the posterior probability distribution:

***

**Algorithm 1** Exact rejection sampler

* Given $N$ the number of simulations
* **for** $i=1$ to $N$ **do**
  + Generate $p'\sim\pi(p)$
  + Simulate $D'\sim f(p')$
  + **if** $D'=D$ **then**
    - Accept $p'$
  + **end if**
* **end for** 
* **return** accepted $p'$ 

***

### Exercise 3

Obtain a sample from the posterior probability distribution with the rejection algorithm:

```{r exact_rejection_sampling, exercise=TRUE, exercise.lines=8}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- _____ # sample from prior distribution (use a beta)
d_prime <- _____ # simulate coin tosses (with binomial)
p_prime_accept <- _____ # keep the values that are equal to the observed data
p_prime_accept
```

```{r exact_rejection_sampling-hint-1}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- rbeta(num_of_sim,a,b)
d_prime <- _____ # simulate coin tosses (with binomial)
p_prime_accept <- _____ # keep the values that are equal to the observed data
p_prime_accept
```

```{r exact_rejection_sampling-hint-2}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- rbeta(num_of_sim,a,b)
d_prime <- rbinom(n    = num_of_sim,
                  size = h+t,
                  prob = p_prime)
p_prime_accept <- _____ # keep the values that are equal to the observed data
p_prime_accept
```


```{r exact_rejection_sampling-solution}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- rbeta(num_of_sim,a,b)
d_prime <- rbinom(n    = num_of_sim,
                  size = h+t,
                  prob = p_prime)
p_prime_accept <- p_prime[which(d_prime==h)]
p_prime_accept
```

### Exercise 4

Use `flip_coin_posterior_approx()` function (see source code for details) to crate a sample from the posterior probability distribution. Create a scatter plot to show the parameter value $p'$ and data $D'$ for all simulations. Highlight simulations that match observed data $D$. Then, use the sample of the posterior probability to plot an estimate of the posterior probability (hint: use `hist()`).

```{r flip_coin_posterior_approx}
flip_coin_posterior_approx <- function(h,t,a,b,num_of_sim){
  p_prime <- rbeta(num_of_sim,a,b)
  d_prime <- rbinom(num_of_sim,h+t,p_prime)
  p_prime_accept <- p_prime[which(d_prime==h)]
  d_prime_accept <- d_prime[which(d_prime==h)]
  return(list(p_prime=p_prime,
              d_prime=d_prime,
              p_prime_accept=p_prime_accept,
              d_prime_accept=d_prime_accept))
}
```


```{r plot_posterior_approx, exercise=TRUE, exercise.lines=20, exercise.setup="flip_coin_posterior_approx"}
h=4;t=6;a=1;b=1;num_of_sim=1000
res <- flip_coin_posterior_approx(h,t,a,b,num_of_sim)

```

```{r plot_posterior_approx-hint}
h=4;t=6;a=1;b=1;num_of_sim=1000
res <- flip_coin_posterior_approx(h,t,a,b,num_of_sim)
plot(res$p_prime,res$d_prime,
     xlab = expression(italic(p)*"'"),
     ylab = expression(italic(D)*"'"))
abline(h = h, col = Reddish_Purple)
points(res$p_prime_accept, res$d_prime_accept, col = Vermillion)

```



```{r plot_posterior_approx-solution}
h=4;t=6;a=1;b=1;num_of_sim=1000
res <- flip_coin_posterior_approx(h,t,a,b,num_of_sim)
plot(res$p_prime,res$d_prime,
     xlab = expression(italic(p)*"'"),
     ylab = expression(italic(D)*"'"))
abline(h = h, col = Reddish_Purple)
points(res$p_prime_accept, res$d_prime_accept, col = Vermillion)
hist(res$p_prime,
     breaks = seq(0,1,0.02),
     col = "grey",
     freq = FALSE,
     ylim = c(0,5),
     main = "", xlab = expression(italic(p)),
     ylab = "probability density")
hist(res$p_prime_accept,
     breaks = seq(0,1,0.02),
     col = Vermillion_transparency,
     freq = FALSE, add = TRUE); box()
lines(seq(0,1,0.001),dbeta(x=seq(0,1,0.001),a+h,b+t),
      col = Vermillion, lwd = 3)
```


### Exercise 5

Get a point estimate and 

```{r coin_toss_credibility_interval, exercise=TRUE, exercise.setup="flip_coin_posterior_approx"}
h=4;t=6;a=1;b=1;num_of_sim=1000
res <- flip_coin_posterior_approx(h,t,a,b,num_of_sim)
p_hat  <- _____
p_95CI <- _____
p_hat; p_95CI
```

```{r coin_toss_credibility_interval-hint}
h=4;t=6;a=1;b=1;num_of_sim=1000
res <- flip_coin_posterior_approx(h,t,a,b,num_of_sim)
p_hat  <- _____(res$p_prime_accept)
p_95CI <- quantile(_____,probs=c(0.025,0.975))
p_hat; p_95CI
```

```{r coin_toss_credibility_interval-solution}
h=4;t=6;a=1;b=1;num_of_sim=1000
res <- flip_coin_posterior_approx(h,t,a,b,num_of_sim)
p_hat  <- median(res$p_prime_accept)
p_95CI <- quantile(res$p_prime_accept,probs=c(0.025,0.975))
p_hat; p_95CI
```

For data $h=4$ and $t=6$, and prior $\mathrm{Beta}(\alpha=1,\beta=1)$ the median is `r qbeta(0.5,1+4,1+6)` and the 95% credibility interval is (`r qbeta(c(0.025,0.975),1+4,1+6)`).


## Approximation 2

We are going to consider now population genetics data. We have two data sets of mitochondrial DNA sequences for 50 individuals of two different hermaphrodite species (*actually, it is fake data*). The data is in two files named `dataset1.txt` and `dataset2.txt` in ms format.

### Exercise 6

Explore the data. How many polymorphic sites are there?

```{r explore_genetic_data, exercise=TRUE}
dataset1 <- read.ms.output(file="/change/to/your/path/dataset1.txt")
dataset2 <- read.ms.output(file="/home/miguel/Work/Teaching/ABCRF tutorial/dataset2.txt")
```







## Approximation 3

