---
#############################################################
#                                                           #
# Click on "Run Document" in RStudio to run this worksheet. #
#                                                           #
#############################################################
title: "Aproximate Bayesian Computation"
author: "Miguel de Navascués"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
```

## Setup

[Colorblind friendly palette](http://jfly.iam.u-tokyo.ac.jp/color/):

```{r color_setup, include=TRUE}
Orange         <- "#E69F00"
Sky_Blue       <- "#56B4E9" 
Bluish_Green   <- "#009E73"
Yellow         <- "#F0E442"
Blue           <- "#0072B2"
Vermillion     <- "#D55E00"
Reddish_Purple <- "#CC79A7"
```

## Approximation 1

Approximate Bayesian Computation belongs is a likelihood-free model-based inference approach. The idea behind likelihood-free methods is that instead of calculating the likelihood is substituted by estimation of the likelihood based on simulations.

### Exercise 1

For the coin toss experiment, estimate the likelihood of $p=0.5$ for $h=4$ and $t=6$ using simulations:

```{r simulation_coin_toss_rbinom, exercise=TRUE}
h = 4; t = 6; p = 0.5
n <- h+t
num_of_sim = _____
h_prime <- rbinom(n=num_of_sim,size=n,prob=p)
_____
```

```{r simulation_coin_toss_rbinom-hint}
h = 4; t = 6; p = 0.5
n <- h+t
num_of_sim = 100
h_prime <- rbinom(n=num_of_sim,size=n,prob=p)
_____(h_prime==h)/_____
```

```{r simulation_coin_toss_rbinom-solution}
h = 4; t = 6; p = 0.5
n <- h+t
num_of_sim = 100
h_prime <- rbinom(n=num_of_sim,size=n,prob=p)
sum(h_prime==h)/num_of_sim
```

The true value (for $p=0.5$, $h=4$ and $t=6$) is `r choose(10, 4) * 0.5^10`. Is the estimate reasonable? How can the estimate be improved?


### Exercise 2

Plot the likelihood profile from using simulation-based estimates of the likelihood. Compare the results with the analytical solution. Why should we bother about a (slower, less accurate) simulation-based solution?

```{r eval=TRUE, echo=TRUE}
flip_coin_likelihood_true <- function(p,h,t){
  likelihood <- choose(h+t,h) * p^h * (1-p)^t
  return (list(p=p,l=likelihood))
}
```


```{r likelihood_function, exercise=TRUE, exercise.lines=17}
flip_coin_likelihood_approx <- function(p,h,t,num_of_sim){
  likelihood <- array(dim=length(p))
  for (i in seq_along(p)){
    likelihood[i] <- _____  # simulate with binomial distribution
  }
  return (list(p=p,l=likelihood))
}
h = 4; t = 6; num_of_sim = 1000
p = seq(0,1,0.01)
l_profile_true   <- flip_coin_likelihood_true(p,h,t)
l_profile_approx <- flip_coin_likelihood_approx(p,h,t,num_of_sim)
plot(x = l_profile_approx$p,
     y = l_profile_approx$l,
     xlab = "p", ylab = "Likelihood", type = "l", col = Bluish_Green, lwd=2)
lines(x = l_profile_true$p,
      y = l_profile_true$l,
      col = Bluish_Green)
```

```{r likelihood_function-solution}
flip_coin_likelihood_approx <- function(p,h,t,num_of_sim){
  likelihood <- array(dim=length(p))
  for (i in seq_along(p)){
    likelihood[i] <- sum(rbinom(n=num_of_sim,size=h+t,prob=p[i])==h)/num_of_sim  
  }
  return (list(p=p,l=likelihood))
}
h = 4; t = 6; num_of_sim = 1000
p = seq(0,1,0.01)
l_profile_true   <- flip_coin_likelihood_true(p,h,t)
l_profile_approx <- flip_coin_likelihood_approx(p,h,t,num_of_sim)
plot(x = l_profile_approx$p,
     y = l_profile_approx$l,
     xlab = "p", ylab = "Likelihood", type = "l", col = Bluish_Green, lwd=2)
lines(x = l_profile_true$p,
      y = l_profile_true$l,
      col = Bluish_Green)
```

Posterior probabilities can be also estimated through the use of simulations using the **rejection algorithm**. The following procedure produces a sample from the posterior probability distribution:

***

**Algorithm 1** Exact rejection sampler

* Given $N$ the number of simulations
* **for** $i=1$ to $N$ **do**
  + Generate $p'\sim\pi(p)$
  + Simulate $D'\sim f(p')$
  + **if** $D'=D$ **then**
    - Accept $p'$
  + **end if**
* **end for** 
* **return** accepted $p'$ 

***

### Exercise 3

Obtain a sample from the posterior probability distribution with the rejection algorithm:

```{r exact_rejection_sampling, exercise=TRUE, exercise.lines=8}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- _____ # sample from prior distribution (use a beta)
d_prime <- _____ # simulate coin tosses (with binomial)
p_prime_accept <- _____ # keep the values that are equal to the observed data
p_prime_accept
```

```{r exact_rejection_sampling-hint-1}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- rbeta(num_of_sim,a,b)
d_prime <- _____ # simulate coin tosses (with binomial)
p_prime_accept <- _____ # keep the values that are equal to the observed data
p_prime_accept
```

```{r exact_rejection_sampling-hint-2}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- rbeta(num_of_sim,a,b)
d_prime <- rbinom(n    = num_of_sim,
                  size = h+t,
                  prob = p_prime)
p_prime_accept <- _____ # keep the values that are equal to the observed data
p_prime_accept
```


```{r exact_rejection_sampling-solution}
h=4; t=6
a=1; b=1; num_of_sim=1000
p_prime <- rbeta(num_of_sim,a,b)
d_prime <- rbinom(n    = num_of_sim,
                  size = h+t,
                  prob = p_prime)
p_prime_accept <- p_prime[which(d_prime==h)]
p_prime_accept
```





## Approximation 2






## Approximation 3

