---
######################################
# Click on "Run Document" in RStudio #
######################################
title: "Good Practices in ABC"
author: "Miguel de Navascués"
output: learnr::tutorial
runtime: shiny_prerendered
bibliography: references.bib
---

This course is available at [Zenodo](http://doi.org/10.5281/zenodo.1435503).



## Setup (just for information)

```{r setup, include=TRUE}
library(learnr)
library(phyclust, quietly=T)
library(abc, quietly=T) 
library(weights, quietly=T)
```

## Compare prior and posterior

Priors can be used to incorporate previous information
on the parameters. In the case of our population genetics example, we could
imagine to have some information on mutation rates (e.g. from pedigrees) and
population sizes (from census). Since $\theta = 2N\mu$, we could draw parameters $N$
and $\mu$ from prior probability distributions and combine them to get the prior of
$\theta$. In order to save time I have generated a reference table from such prior with the following code:

```{r eval=FALSE, echo=TRUE}
sample_size = 50
num_of_sim = 1000000

mu_prime <- 10^runif(num_of_sim,min=-7.5,max=-6)
N_prime <- 10^runif(num_of_sim,min=6,max=7.5)
theta_prime <- 2*N_prime*mu_prime

msout <- ms(nsam = sample_size,
            opt = "-t tbs",
            tbs.matrix = cbind(theta_prime))
msout <- read.ms.output(txt=msout)
S_prime  <- S(msout)
PI_prime <- PI(msout)
NH_prime <- NH(msout)
TD_prime <- TajimaD(msout)
ref_table <- data.frame(theta=theta_prime,
                        S=S_prime,
                        PI=PI_prime,
                        NH=NH_prime,
                        TD=TD_prime)
saveRDS(ref_table,file="ref_table_1b.RDS")
```




### Exercise 1

```{r unexpected_prior, exercise=TRUE, exercise.lines=30}
target <- data.frame(S  = c(50,23),
                     PI = c(11.56163,1.354286),
                     NH = c(21,18),
                     TD = c(0.123358,-2.386268),
                     row.names = c("octomanati","rinocaracol"))
ref_table <- readRDS(file="/home/miguel/Work/Teaching/ABCRF tutorial/ref_table_1b.RDS")

num_of_sims <- 100000 # use a subset of simulation from a total of 1 000 000

SS          <- target[1,"TD"]
SS_prime    <- ref_table[seq_len(num_of_sims),"TD"]
theta_prime <- ref_table$theta[seq_len(num_of_sims)]

tolerance <- 0.05
abc_result <- abc(target = SS,
                  param = theta_prime,
                  sumstat = SS_prime,
                  tol = tolerance,
                  method = "rejection")

theta_prime_accept     <- as.vector(abc_result$unadj.values)
hist(x = log10(theta_prime_accept),
     breaks = seq(-2,3,0.05),
     col = "blue",
     freq = FALSE,
     ylim = c(0,1), xlim=c(-1,2),
     xlab = expression(log(theta)),main="",
     ylab = "probability density")
```

```{r unexpected_prior-solution}
target <- data.frame(S  = c(50,23),
                     PI = c(11.56163,1.354286),
                     NH = c(21,18),
                     TD = c(0.123358,-2.386268),
                     row.names = c("octomanati","rinocaracol"))
ref_table <- readRDS(file="/home/miguel/Work/Teaching/ABCRF tutorial/ref_table_1b.RDS")

num_of_sims <- 100000 # use a subset of simulation from a total of 1 000 000

SS          <- target[1,"TD"]
SS_prime    <- ref_table[seq_len(num_of_sims),"TD"]
theta_prime <- ref_table$theta[seq_len(num_of_sims)]

tolerance <- 0.05
abc_result <- abc(target = SS,
                  param = theta_prime,
                  sumstat = SS_prime,
                  tol = tolerance,
                  method = "rejection")

theta_prime_accept     <- as.vector(abc_result$unadj.values)
hist(x = log10(theta_prime_accept),
     breaks = seq(-2,3,0.05),
     col = "blue",
     freq = FALSE,
     ylim = c(0,1), xlim=c(-1,2),
     xlab = expression(log(theta)),,main="",
     ylab = "probability density")
hist(x = log10(theta_prime),
     breaks = seq(-2,3,0.05),
     col = "grey", freq=FALSE, add=TRUE )
```

